PREPRINT -- Discriminating Seagrasses From Green Macroalgae in European Intertidal areas using high resolution multispectral drone imagery -- PREPRINT.

### 2024-04-29 docs/_tex/index.tex
pierre's comments
add pierre's comment

\begin{abstract}
Coastal areas support seagrass meadows, which offer crucial ecosystem
services including erosion control and carbon sequestration. However,
these areas are increasingly impacted by human activities, leading to
seagrass decline and habitat fragmentation. In situ surveys,
traditionally performed to monitor these ecosystems face limitations on
temporal and spatial coverage, particularly in intertidal zones,
prompting the use of satellite data within monitoring programs. Yet,
satellite remote sensing struggles with spatial and spectral resolution,
making it difficult to discriminate seagrass from other macrophytes in
highly heterogenous meadows. Drone images at a very high spatial
resolution offer a promising solution to address challenges related to
spatial heterogeneity and intrapixel mixture. This study focuses on
using drone acquisitions with a ten spectral band sensor mirroring those
of Sentinel-2, for mapping intertidal macrophytes and effectively
discriminating between seagrass and green macroalgae. Nine drone flights
were conducted at two different altitudes (12m and 120m) across
heterogeneous intertidal European habitats in France and Portugal. Low
altitude flights were used to train a Deep Learning classifier based on
Neural Networks to discrimintate among five taxonomic classes of
intertidal vegetation: Magnoliopsida (Seagrass), Chlorophyceae (Green
macroalgae), Phaeophyceae (Brown algae), Rhodophyceae (Red macroalgae)
and benthic Bacillariophyceae (Diatoms). Classification of drone imagery
resulted in an overall accuracy of 94\% across all the sites and images,
covering a total area of 467 000 m². The model exhibited an accuracy of
96.4\% in identifying seagrass. This work showed that although they
share the same pigmentary composition, seagrass and green algae can be
discriminated using a multispectral sensor with ten spectral bands at
444, 475, 531, 560, 650, 668, 705, 717, 740 and 840 nm. The algorithm
should now be adapted for Sentinel-2 to see if this discrimination is
still operational at a coarser spatial resolution.
\end{abstract}

@@

While satellite remote sensing (RS) provides temporally consistent
observations over large spatial scales, its utilization over intertidal
areas is limited by several constraints. Satellite missions with a high
temporal resolution (e.g.~daily MODIS observation) are limited by a
coarse spatial resolution (\textgreater100m) over these restricted
extension areas. Missions with a high spatial resolution such as
Sentinel-2 (10m) or Landsat8/9 (30m) can be limited by low spectral
resolution. The limited number of spectral bands is indeed a challenge
to accurately discriminate seagrass from others co-existing macrophytes.
In particular Chlorophyceae (green algae) and marine Magnoliopsida
(seagrass) share the same pigment composition \citetext{\citealp[
]{ralph2002}; \citealp{Douay2022}}. Therefore, to someone not
specialized in the field, their spectral signatures may appear to be
alike \citetext{\citealp[ ]{Davies2023}; \citealp{bannari2022}}.
Recently, using advanced machine-learning algorithms trained with a
large hyperspectral library of more than 300 field reflectance spectra,
\citep{Davies2023} demonstrated that it was possible to discriminate
Magnoliopsida from Chlorophyceae using radiometric data acquired at
Sentinel-2 's spectral resolution. However the application of this
approach to satellite RS remains to be validated. Moreover patches of
green algae can develop at small spatial scales that are not observable
using non-commercial satellite imagery \citep{tuya2013}, especially
during the initial stage of a green tide.

Drones can potentially fill the gaps left by satellite RS and in situ
measurements, due to their ability to provide spatially-explicit
observations at very high spatial resolutions (pixel size from cm to mm)
while capturing data at multi-spectral resolution \citetext{\citealp[
]{fairley2022drone}; \citealp{oh2017use}}. The versatility of drones
allows for their application across a diverse thematic range , from
coastal zone management \citetext{\citealp[ ]{adade2021}; \citealp[
]{casella2020}; \citealp{angnuureng2022}} to mapping species
distribution \citetext{\citealp[ ]{joyce2023}; \citealp[
]{tallam2023}; \citealp[ ]{Roca2022}; \citealp[
]{Roman2021}; \citealp{Brunier2022Topographic}}. However, when applied
to coastal habitat mapping, most case studies are limited to a single
flight, restricting the generalizability of their application over wider
geographical scales \citetext{\citealp[ ]{Roman2021}; \citealp[
]{collin2019improving}; \citealp[
]{rossiter2020uav}; \citealp{Brunier2022Topographic}}. The present study
aimed at analyzing the potential of multispectral drone RS to map
intertidal macrophytes, with a particular focus on discriminating
Magnoliopsida and Chlorophyceae (Seagrass and Green Algae respectively).
Ten drone flights were performed over soft-bottom intertidal areas along
two European countries (France and Portugal), covering a wide range of
habitats, from monospecific seagrass meadows to meadows mixed with
green, or red macroalgae. A deep learning algorithm was trained and
validated for macrophyte discrimination, emphasizing applicability
across diverse sites without a loss of prediction accuracy.

@@

\caption{\label{fig-map}Location of drone flights in France and
Portugal. A: Gulf of Morbihan (Two sites), B: Bourngeuf Bay (Two sites),
C: Ria de Aveiro Coastal Lagoon (Three sites). Green represents
intertidal areas.}

@@

The nine drone flights were utilized to produce a total of nine
prediction maps, each of them associated with a probability map that
indicates the probability of the selected class for every pixel. The
low-altitude flight conducted in Gafanha, Portugal, represents the site
with the highest complexity (Figure~\ref{fig-GafLow}). Among the five
vegetation classes on which the model was trained, four were present on
this site, with Chlorophyceae and Rhodophyceae mixed with the seagrass
meadow. There was also Bacillariophyceae forming biofilms on parts of
sediment surface. Although the seagrass bed was solely composed of
\emph{Nanozostera noltei}, various colors of this specie could be
observed from dark green (corresponding to healthy leaves) to
whitish/brown (when leaves were discolored having lost their
pigmentation). Regardless of color, Magnoliopsida pixels were accurately
predicted by the model.

@@

The high-altitude flight over Gafanha covered a total area of
approwimately 1 km² (Figure~\ref{fig-GafHigh}). A channel contouring a
small island was masked in the prediction map. Most of the intertidal
area was classified as Magnoliopsida by the model, including seagrass
patches with discolored leaves. Only a few pixels were classified as
Chlorophyceae at this spatial scale. Furthermore, the area that was
classified as Bacillariophyceae in the low-altitude flight remained
mostly classified as such in the high-altitude flight, though some
pixels were classified as Magnoliopsida. Patches of Rhodophyceae were
correctly classified. In the northern part of the site and near the land
eadges, patches of the schorre angiosperm \emph{Spartina sp.} were
misclassified, either as Magnoliopsida or as Phaeophyceae.

@@

@@

In this study, a key innovation lies in the utilization of drone flights
at two different altitudes (12m and 120m) for constructing the neural
network model. The lower altitude flights (8mm spatial resolution)
enabled precise selection of pure pixels to train the neural network
model. This methodology implies a consistency between the reflectance of
both altitudes. Figure~\ref{fig-CompareRef} depicts the relationship
between reflectance from a low and a high-altitude flights conducted
over the same area. Overall there was a good agreement between the two
altitudes. There was however a slight underestimation for raw
reflectance values in the high-altitude flight, particularly for higher
reflectance values (Figure~\ref{fig-CompareRef} A). Since both flights
were conducted over vegetation areas, the highest reflectance values
correspond to the infrared part of the spectrum. This different is not
present anymore when reflectance values of both flights have been
standardized (Equation~\ref{eq-std} ; Figure~\ref{fig-CompareRef} B).

@@

@@

A total of 536,000 pixels were used to validate the Neural Network
classifier. The sites with the lowest and highest number of validation
data were Gafanha Low (17316 pixels) and Marinha Lanzarote (159713
pixels), respectively. Model global accuracy was 94.26\% with a Kappa
coefficient of 0.92 (Figure~\ref{fig-Validation}). The lowest performing
site was Gafanha High (global accuracy of 75.45\%) whereas Mataducos was
the site with the most accurate prediction (global accuracy of 98.05\%).
Overall, the classes Phaeophyceae, Magnoliopsida, Sediment and
Rhodophyceae were correctly classified with a balanced accuracy of 1,
0.96, 0.96 and 0.91 respectively. Bacillariophyceae was the least
performing class (accuracy of 0.72 ) mainly due to the confusion with
Magnoliopsida and Sediment.

@@

@@

The computation of the variable importance made it possible to identify
which bands were the most useful for class prediction
(Figure~\ref{fig-VIP}). The spectral bands at 444, 717 and 842 nm of the
Micasense camera did not provide as important information to
discriminate any of the vegetation classes. The band at 531 nm was the
only important predictor for the classifier to accurately predict
Chlorophyceae. In fact, at this wavelength, the Chlorophyceae spectra
showed the highest reflectance among of all vegetation classes. The
bands at 531 and 740 nm were the most important predictors for
Phaeophyceae, corresponding to the lowest reflectance among all classes.
Bands at 475 and 560 nm were the most important predictors for
Bacillariophyceae and Rhodophyceae, respectively. Four predictors,
ranging from the green (560 nm) to the RedEdge (705 nm) bands were
important to accurately predict Magnoliopsida.

\subsection{Effect of the flight height on the
prediction}\label{effect-of-the-flight-height-on-the-prediction}

@@

Using the very high resolution low altitude flight (8 mm pixels), we
determined the minimal percent cover required to correctly classify a
given class within the corresponding high altitude flight
(Figure~\ref{fig-upscaling}). When the percent cover of a given class
was 100 \%, coarser high-flight pixels are well classified for all the
classes except for Bare Sediment, which was only well classified 80\% of
the time. This phenomenon is likely attributed to the time gap between
the two flights, allowing for microphytobenthos migration to surface
during low tide, consequently altering the model's classification from
bare sediment to Bacillariophyceae. A percent cover of at least 80\% was
sufficient to have all the pixels of high altitude flights correctly
classified, with the exception of Magnoliopsida that required a higher
percent cover (\textgreater90 \%) to be well classified. Concerning the
probability of each class, the highest percent cover was needed to
confidently predict Bacillariophyceae. To predict Chlorophyceae with a
confidence of 0.85, a percent cover of 93 \% was needed, 90 \% for
Magnoliopsida, 92 \% for Rhodophyceae and 97 \% for Bacillariophyceae.

@@

@@

\caption{\label{fig-Pigm}Photosynthetic and carotenoid pigments present
(Green) or absent (Red) in each taxonomic class present in the Neural
Network Classifier, along with their absorption wavelength measured with
spectroradiometer. Chl b: chlorophyll b, Chl c: chlorophyll c, Fuco:
fucoxanthin, Zea: zeaxanthin, Diad: diadinoxanthin, Lut: lutein, Neo:
neoxanthin, PE: phycoerythrin, PC: phycocyanin.}

@@

The primary objective of this study was to develop a method for the
accurate classification of macrophytes on intertidal mudflats,
specifically focusing on distinguishing between Chlorophyceae (green
macroalgae) and marine Magnoliopsida (seagrasses) using multispectral
drone data. The ability to differentiate between various types of
vegetation, as demonstrated in this study, plays a critical role in
ecological monitoring and coastal management. By distinguishing between
seagrasses and macroalgae, our approach facilitates targeted
conservation strategies, enabling more effective preservation and
restoration efforts in coastal ecosystems. The discrimination of
seagrasses from green macroalgae presents significant challenges
\citetext{\citealp[ ]{oiry2021using}; \citealp[
]{bannari2022}; \citealp{veettil2020opportunities}}, primarily due to
the similarities on their spectal signatures in the visible region of
the eletromagnetic spectra due to shared pigment
, Figure~\ref{fig-Pigm}, but also by the frequent spatial mixing of
these macrophytes. Besides chlorophyll-a that is common to all
vegetation types, these two green macrophytes have in common other
photosynthetic pigments such as chlorophyll-b, and accessory carotenoid
pigments such as zeaxantin, lutein and neoxanthin
(Figure~\ref{fig-Pigm}). This spectral overlap is further complicated by
the intermingling of seagrasses and green macroalgae in the same spatial
locations, posing a significant challenge for remote sensing aimed at
accurately mapping and monitoring coastal ecosystems. Our study
addresses these complexities by using high-resolution drone imagery with
10 spectral bands. \citep{Davies2023} have shown that having at least
eight spectral bands ranging between 500 nm to 850 nm including a green
band at 530 nm and another one in the RedEdge region at 730 nm was
crucial to accurately discriminate amoung green macrophytes.

@@

Meeting these two criteria, the Micasense RedEdge-MX DUAL camera used in
this study, enabled the classifier to achieve 97\% of accuracy between
these two classes (Figure~\ref{fig-ValidationGreen}). Even if the
pigment composition of green macrophytes is similar, differences in the
spectral shape can still be observed (Figure~\ref{fig-vegetation}).
Several factors can explain these differences such has different
concentration and/or pigment ratios \citep{bargain2013seasonal},
different cellular organisation and the disposition of the plant over
the sediment surface \citetext{\citealp[ ]{beach1997vivo}; \citealp[
]{kirk1994light}; \citealp{hedley2018influence}}.

The VIP analysis of the Neural Network model (Figure~\ref{fig-VIP})
identified that the band at 531 nm was the most important one for
accurately identifying Chlorophyceae. In fact, at this wavelength,
Chlorophyceae exhibited the highest reflectance (raw and standardised)
among all other classes, highlighting the difference in accessory
pigments (i.e.~carotenoid) between seagrasses and green macroalgae
\citep{repolho2017seagrass}.

Concerning Phaeophyceae, the thick cell walls of these macroalgae
\citep{charrier2021growth} make it more reflective in the infrared part
of the spectra \citep{Slaton2001} whereas the presence of Fucoxanthin
and Zeaxanthin result in a low reflectance in the visible region
(Figure~\ref{fig-Pigm} ; Figure~\ref{fig-VIP}). These two key features
have been identified by the Neural Network as the two principal
predictors to accurately identify brown algae (Figure~\ref{fig-VIP}).
Similarly, the presence of phycoerythrin and phycocyanin in Rhodophyceae
contributes to the lowest reflectance among all classes in the spectral
range of 560 to 615 nm (Figure~\ref{fig-VIP}). Indeed the band at 560 nm
has been identified as important for identifying this class likely due
to phycoerythrin absorption at this wavelength (\textbf{REF}). Regarding
Bacillariophyceae, the VIP analysis (Figure~\ref{fig-VIP}) indicated
that 475 nm was the most important predictor for this class. This is not
a usual diagnostic wavelength for Bacillariophyceae which is mainly
identified by the presence of fucoxanthin and chlorophyll c
(Figure~\ref{fig-Pigm}), but it correspond to the absorption of
diadinoxanthin an accessory pigment characteristic of this class
\citep{meleder2003spectrometric}. Furthermore, it is the vegetation with
the lowest concentration of chlorphyll-a, pigment absorbing light both
in the blue and the red. The transparency of Bacillariophyceae makes the
reflectance of the sediment part of the overall reflectance of
Bacillariophyceae, further explaining the high reflectance in the blue.

\subsection{Spectral Spatial Temporal Resolution impact on the
prediction}\label{spectral-spatial-temporal-resolution-impact-on-the-prediction}

@@

\subsection{Application of the model}\label{application-of-the-model}

Climate change, global warming, alien and invasive species development,
coastal erosion, and sealevel rise are expected to continue impacting
coastal ecosystems in the future
\citetext{\citealp{SCHIBALSKI2022101414}; \citealp[
]{holon2018predictive}; \citealp{marquet2024global}}. Therefore the
demand for meaningful and efficient monitoring methods for these
habitats has never been greater \citetext{\citealp[
]{muller2018satellite}; \citealp[
]{villalobos2023remote}; \citealp{oiry2021using}}. Our findings,
particularly the improved discrimination of seagrass from other
intertidal vegetation classes, underscore the potential of drone-based
remote sensing to support diverse applications, from biodiversity
conservation to climate change adaptation strategies.

With the increase in nutrient concentrations in coastal waters,
macroalgal blooms are becoming increasingly common in many regions
around the world \citetext{\citealp[
]{sutton2011european}; \citealp{ye2011green}}. These blooms can have
negative impacts on local economic activities, including human health,
fishing and aquaculture, tourism, and recreational activities
\citetext{\citealp[ ]{villares1999nitrogen}; \citealp{ye2011green}}. The
first green tide events (i.e.~bloom of green macroalgae of the genus
\emph{Ulva}) were reported in Brittany, France, back in the 1970s and
have since been a concern for local policies and economic activities
\citep{menesguen2018marees}. Some regions of the world have witnessed an
increase in brown macroalgae blooms, predominantly involving algae of
the genus \emph{Sargassum} \citep{louime2017sargassum}. Satellite remote
sensing has proven to be a valuable tool for mapping the spatial and
temporal extent of macroalgal blooms worldwide. However, due to its
limitations in spatial resolution, it can only effectively map
well-developed blooms \citetext{\citealp[
]{rs13081408}; \citealp{klemas2012remote}}. High spatial resolution
drone imagery, coupled with an accurate classification algorithm, could
be utilized to map the early stages of macroalgal blooms in areas known
to have regular blooms or in new sites. Indeed, this approach could
provide early warning alerts to local policymakers.

Employing traditional sampling methods to monitor these ecosystems is
time and resource-intensive, and the findings are often difficult to
up-scale. Earth observation methods can bridge this gap and meet the
needs for monitoring coastal ecosystems
\citep{papathanasopoulou2019satellite}. The retrieval of Essential
Biodiversity Variables and Essential Ocean Variables through satellite
observations is increasingly common, enabling comprehensive monitoring
of entire ecosystems over extended time periods \citetext{\citealp[
]{ratnarajah2023monitoring}; \citealp{ZOFFOLI2020112020}}. The
significance of monitoring the coastal environment, particularly in
light of the Water Framework Directive (WFD ; \citep{WFD2000}), cannot
be overstated in environmental management and conservation efforts. The
WFD mandates the achievement and maintenance of ``good ecological
status'' for all European waters, which necessitates a comprehensive
understanding and monitoring of aquatic ecosystems, including coastal
habitats like seagrass beds \citetext{\citealp[
]{foden2007angiosperms}; \citealp[
]{nordlund2024one}; \citealp{Zoffoli2021}}.

>>>>

\begin{abstract}
Coastal areas support seagrass meadows, which offer crucial ecosystem
services including erosion control and carbon sequestration. However,
these areas are increasingly impacted by human activities, leading to
seagrass decline and habitat fragmentation. In situ surveys,
traditionally performed to monitor these ecosystems face limitations on
temporal and spatial coverage, particularly in intertidal zones,
prompting the use of satellite data within monitoring programs. Yet,
satellite remote sensing struggles with spatial and spectral resolution,
making it difficult to discriminate seagrass from other macrophytes in
highly heterogenous meadows. Drone images at a very high spatial
resolution offer a promising solution to address challenges related to
spatial heterogeneity and intrapixel mixture. This study focuses on
using drone acquisitions with a ten spectral band sensor mirroring those
of Sentinel-2, for mapping intertidal macrophytes and effectively
discriminating between seagrass and green macroalgae. Nine drone flights
were conducted at two different altitudes (12m and 120m) across
heterogeneous intertidal European habitats in France and Portugal. Low
altitude flights were used to train a Deep Learning classifier based on
Neural Networks to discrimintate among five taxonomic classes of
intertidal vegetation: Magnoliopsida (Seagrass), Chlorophyceae (Green
macroalgae), Phaeophyceae (Brown algae), Rhodophyceae (Red macroalgae)
and benthic Bacillariophyceae (Diatoms). Classification of drone imagery
resulted in an overall accuracy of 94\% across all the sites and images,
covering a total area of 467 000 m². The model exhibited an accuracy of
96.4\% in identifying seagrass. Importantly, seagrass and green algae
can be discriminated, although they share the same pigment composition.
As the algorithm was developed for a multispectral camera with ten
spectral bands in the visible and near-infrared, it could be adapted to
the Multi-Spectral Instrument (MSI) onboard Sentinel-2 thus offering
promising perspectives for satellite remote sensing of intertidal
biodiversity over lager scales.
\end{abstract}

@@

While satellite remote sensing (RS) provides temporally consistent
observations over large spatial scales, its utilization over intertidal
areas is limited by several constraints. Satellite missions with a high
temporal resolution (e.g.~daily MODIS observation) are limited by a too
coarse spatial resolution (\textgreater100m) to accurately map patchy
seagrass meadows. Missions with a high spatial resolution such as
Sentinel-2 (10m) or Landsat8/9 (30m) can be limited by low spectral
resolution. The limited number of spectral bands is indeed a challenge
to accurately discriminate seagrass from others co-existing macrophytes.
In particular Chlorophyceae (green algae) and marine Magnoliopsida
(seagrass) share the same pigment composition \citetext{\citealp[
]{ralph2002}; \citealp{Douay2022}}. Therefore, to someone not
specialized in the field, their spectral signatures may appear to be
alike \citetext{\citealp[ ]{Davies2023}; \citealp{bannari2022}}.
Recently, using advanced machine-learning algorithms trained with a
large hyperspectral library of more than 300 field reflectance spectra,
\citep{Davies2023} demonstrated that it was possible to discriminate
Magnoliopsida from Chlorophyceae using radiometric data acquired at
Sentinel-2 's spectral resolution. However the application of this
approach to satellite RS remains to be validated. Moreover patches of
green algae can develop at small spatial scales that are not observable
using non-commercial satellite imagery \citep{tuya2013}, especially
during the initial stage of a green tide.

Drones can potentially fill the gaps left by satellite RS and in situ
measurements, due to their ability to provide spatially-explicit
observations at very high spatial resolutions (pixel size from mm to cm)
while capturing data at multi-spectral resolution \citetext{\citealp[
]{fairley2022drone}; \citealp{oh2017use}}. The versatility of drones
allows for their application across a diverse thematic range , from
coastal zone management \citetext{\citealp[ ]{adade2021}; \citealp[
]{casella2020}; \citealp{angnuureng2022}} to mapping species
distribution \citetext{\citealp[ ]{joyce2023}; \citealp[
]{tallam2023}; \citealp[ ]{Roca2022}; \citealp[
]{Roman2021}; \citealp{Brunier2022Topographic}}. However, when applied
to coastal habitat mapping, most case studies were previously limited to
a single flight, restricting the generalizability of their application
over wider geographical scales \citetext{\citealp[
]{Roman2021}; \citealp[ ]{collin2019improving}; \citealp[
]{rossiter2020uav}; \citealp{Brunier2022Topographic}}. The present study
aimed at analyzing the potential of multispectral drone RS to map
intertidal macrophytes over a diverse biogeographical range, with a
particular focus on discriminating Magnoliopsida and Chlorophyceae
(Seagrass and Green Algae respectively). Ten drone flights were
performed over soft-bottom intertidal areas along two European countries
(France and Portugal), covering a wide range of habitats, from
monospecific seagrass meadows to meadows mixed with green, or red
macroalgae. A deep learning algorithm was trained and validated for
macrophyte discrimination, emphasizing applicability across diverse
sites without a loss of prediction accuracy.

@@

\caption{\label{fig-map}Location of drone flights in France and
Portugal. A: Gulf of Morbihan (Two sites), B: Bourngeuf Bay (Two sites),
C: Ria de Aveiro Coastal Lagoon (Three sites). Green areas represents
the intertidal zone.}

@@

Each drone flight was used to produce a prediction map, as well as a
probability map that indicates the likelihood of the selected class for
every pixel. The low-altitude flight conducted in Gafanha, Portugal,
represents the site with the highest complexity
(Figure~\ref{fig-GafLow}). Among the five vegetation classes on which
the model was trained, four were present on this site, with
Chlorophyceae and Rhodophyceae mixed with the seagrass meadow. There was
also Bacillariophyceae forming biofilms on parts of sediment surface.
Although the seagrass bed was solely composed of \emph{Nanozostera
noltei}, various colors of this specie could be observed from dark green
(corresponding to healthy leaves) to whitish/brown (when leaves were
discolored having lost their pigmentation). Regardless of color,
Magnoliopsida pixels were accurately predicted by the model.

@@

The high-altitude flight over Gafanha covered a total area of
\textasciitilde1 km² (Figure~\ref{fig-GafHigh}). A channel contouring a
small island was masked in the prediction map. Most of the intertidal
area was classified as Magnoliopsida by the model, including seagrass
patches with discolored leaves. Only a few pixels were classified as
Chlorophyceae at this spatial scale. Furthermore, the area that was
classified as Bacillariophyceae in the low-altitude flight remained
mostly classified as such in the high-altitude flight, though some
pixels were classified as Magnoliopsida. Patches of Rhodophyceae were
correctly classified. In the northern part of the site and near the land
eadges, patches of the schorre angiosperm \emph{Spartina sp.} were
misclassified, either as Magnoliopsida or as Phaeophyceae.

@@

In this study, a key innovation lies in the utilization of drone flights
at two different altitudes (12m and 120m) for constructing the neural
network model. The lower altitude flights (8mm spatial resolution)
enabled precise selection of pure pixels to train the neural network
model. This methodology implies a consistency between the reflectance
measured from both altitudes. Figure~\ref{fig-CompareRef} depicts the
relationship between reflectance from a low and a high-altitude flights
conducted over the same area.

@@

Overall there was a good agreement between the two altitudes. There was
however a slight underestimation for raw reflectance values in the
high-altitude flight, particularly for higher reflectance values
(Figure~\ref{fig-CompareRef} A). Since both flights were conducted over
vegetation areas, the highest reflectance values correspond to the
infrared part of the spectrum. This different is not present anymore
when reflectance values of both flights have been standardized
(Equation~\ref{eq-std} ; Figure~\ref{fig-CompareRef} B).

@@

A total of 536,000 pixels were used to validate the Neural Network
classifier. The sites with the lowest and highest number of validation
data were Gafanha Low (17316 pixels) and Marinha Lanzarote (159713
pixels), respectively. Model global accuracy was 94.26\% with a Kappa
coefficient of 0.92 (Figure~\ref{fig-Validation}).

@@

The lowest performing site was Gafanha High (global accuracy of 75.45\%)
whereas Mataducos was the site with the most accurate prediction (global
accuracy of 98.05\%). Overall, the classes Phaeophyceae, Magnoliopsida,
Sediment and Rhodophyceae were correctly classified with a balanced
accuracy of 1, 0.96, 0.96 and 0.91 respectively. Bacillariophyceae was
the least performing class (accuracy of 0.72 ) mainly due to the
confusion with Magnoliopsida and Sediment.

@@

The computation of the variable importance made it possible to identify
which bands were the most useful for class prediction
(Figure~\ref{fig-VIP}).

@@

The spectral bands at 444, 717 and 842 nm of the Micasense camera did
not provide as important information to discriminate any of the
vegetation classes. The band at 531 nm was the only important predictor
for the classifier to accurately predict Chlorophyceae. In fact, at this
wavelength, the Chlorophyceae spectra showed the highest reflectance
among of all vegetation classes. The bands at 531 and 740 nm were the
most important predictors for Phaeophyceae, corresponding to the lowest
reflectance among all classes. Bands at 475 and 560 nm were the most
important predictors for Bacillariophyceae and Rhodophyceae,
respectively. Four predictors, ranging from the green (560 nm) to the
RedEdge (705 nm) bands were important to accurately predict
Magnoliopsida.

\subsection{Effect of the percent cover on the
prediction}\label{effect-of-the-percent-cover-on-the-prediction}

Using the very high resolution low altitude flight (8 mm pixels), we
determined the minimal percent cover required to correctly classify a
given class within the corresponding high altitude flight
(Figure~\ref{fig-upscaling}).

@@

When the percent cover of a given class was 100 \%, coarser high-flight
pixels are well classified for all the classes except for Bare Sediment,
which was only well classified 80\% of the time. This phenomenon is
likely attributed to the time gap between the two flights, allowing for
microphytobenthos migration to surface during low tide, consequently
altering the model's classification from bare sediment to
Bacillariophyceae. A percent cover of at least 80\% was sufficient to
have all the pixels of high altitude flights correctly classified, with
the exception of Magnoliopsida that required a higher percent cover
(\textgreater90 \%) to be well classified. Concerning the probability of
each class, the highest percent cover was needed to confidently predict
Bacillariophyceae. To predict Chlorophyceae with a confidence of 0.85, a
percent cover of 93 \% was needed, 90 \% for Magnoliopsida, 92 \% for
Rhodophyceae and 97 \% for Bacillariophyceae.

@@

The primary objective of this study was to develop a method for the
accurate classification of macrophytes on intertidal mudflats,
specifically focusing on distinguishing between Chlorophyceae (green
macroalgae) and marine Magnoliopsida (seagrasses) using multispectral
drone data. The ability to differentiate between various types of
vegetation plays a critical role in ecological monitoring and coastal
management. By distinguishing between seagrasses and macroalgae, our
approach facilitates targeted conservation strategies, enabling more
effective preservation and restoration efforts in coastal ecosystems.
The discrimination of seagrasses from green macroalgae presents two main
challenges \citetext{\citealp[ ]{oiry2021using}; \citealp[
]{bannari2022}; \citealp{veettil2020opportunities}}. First these two
macrophytes have a similar pigment composition: chlorophyll-a (common to
all vegetation types), chlorophyll-b (an additional photosynthetic
pigment), and accessory carotenoids such as zeaxanthin, lutein and
neoxanthin (Figure~\ref{fig-Pigm}). Second, seagrass and green
macroalgae frequently co-occur in intertidal areas, and can intermingle
within a remote sensing pixel if the spatial resolution is too low.
Here, the issue of intra-pixel mixing was resolved thanks to the very
high spatial resolution of the drone, and the risk of spetral confusion
was avoided thanks to the efficiency of the multispectral classifier.
Our results confirm a recent study based on in situ radiometry, where it
was demonstrated that a radiometer with at least eight spectral bands
ranging from 500 to 850 nm and including a green band at 530 nm and a
RedEdge band at 730 nm was crucial to accurately discriminate green
macrophytes \citep{Davies2023}.

@@

\caption{\label{fig-Pigm}Photosynthetic and carotenoid pigments present
(Green) or absent (Red) in each taxonomic class present in the Neural
Network Classifier, along with their absorption wavelength measured with
spectroradiometer. Chl-b: chlorophyll-b, Chl-c: chlorophyll-c, Fuco:
fucoxanthin, Zea: zeaxanthin, Diad: diadinoxanthin, Lut: lutein, Neo:
neoxanthin, PE: phycoerythrin, PC: phycocyanin.}

@@

@@

Meeting these two criteria, the Micasense RedEdge-MX DUAL camera used in
this study, enabled the classifier to achieve 97\% of accuracy between
these two classes (Figure~\ref{fig-ValidationGreen}). Even if the
pigment composition of green macrophytes is similar, differences in the
spectral shape can be observed, with green algae having a higher
reflectance peak at 560 nm as well as a higher NIR plateau than seagrass
(Figure~\ref{fig-vegetation}). Such differences were previously
attributed to differences in pigments concentration and/or ratios
\citep{bargain2013seasonal}, cellular structure as well as in the
disposition of the plant over the sediment surface \citetext{\citealp[
]{beach1997vivo}; \citealp[
]{kirk1994light}; \citealp{hedley2018influence}}.

The VIP analysis (Figure~\ref{fig-VIP}) identified that the band at 531
nm was the most important for accurately identifying Chlorophyceae. In
fact, at this wavelength, Chlorophyceae exhibited the highest
reflectance among all other classes, highlighting the difference in
carotenoids to chlorophyll-a ratios between seagrasses and green
macroalgae \citep{repolho2017seagrass}.

Concerning Phaeophyceae, the thick cell walls of these macroalgae
\citep{charrier2021growth} make it more reflective in the infrared part
of the spectra \citep{Slaton2001} whereas the presence of fucoxanthin
and zeaxanthin result in a low reflectance in the visible region
(Figure~\ref{fig-Pigm} ; Figure~\ref{fig-VIP}). These two key features
have been identified by the Neural Network as the two principal
predictors to accurately identify brown algae (Figure~\ref{fig-VIP}).
Similarly, the presence of phycoerythrin and phycocyanin in Rhodophyceae
contributes to the lowest reflectance among all classes in the spectral
range from 560 to 615 nm (Figure~\ref{fig-VIP}). Indeed the band at 560
nm has been identified as important for identifying this class likely
due to phycoerythrin absorption at this wavelength. Regarding
Bacillariophyceae, the VIP analysis (Figure~\ref{fig-VIP}) indicated
that 475 nm was the most important predictor for this class. Indeed, the
reflectance at 475 nm was higher for Bacillariophyceae than for any
other vegetation class (Figure~\ref{fig-vegetation}), very likely due to
the low biomass (and associated concentration of blue-absorbing
pigments) of microphytobenthos biofilms compared to seagrass and
macroalgae.

\subsection{Altitude and Temporal Effects on Vegetation Prediction
Accuracy}\label{altitude-and-temporal-effects-on-vegetation-prediction-accuracy}

@@

\subsection{Towards climate and biodiversity
applications}\label{towards-climate-and-biodiversity-applications}

Climate change, global warming, eutrophication, alien and invasive
species development, coastal erosion, and sealevel rise are expected to
continue impacting coastal ecosystems in the future
\citetext{\citealp{SCHIBALSKI2022101414}; \citealp[
]{holon2018predictive}; \citealp{marquet2024global}}. and the demand for
meaningful and efficient monitoring of coastal habitats has never been
higher\citetext{\citealp[ ]{muller2018satellite}; \citealp[
]{villalobos2023remote}; \citealp{oiry2021using}}. Our findings,
particularly the improved discrimination of seagrass and green
macroalgae from other intertidal vegetation classes, highlight the
potential of drone-based remote sensing to support diverse applications,
from biodiversity conservation to climate change adaptation strategies.

Because of coastal eutrophication, macroalgal blooms are becoming
increasingly common in many regions around the world \citetext{\citealp[
]{sutton2011european}; \citealp{ye2011green}}. These blooms can have
negative impacts on human health and local economic activities,
including human health, fishing and aquaculture, tourism, and
recreational activities \citetext{\citealp[
]{villares1999nitrogen}; \citealp{ye2011green}}. The first green tide
events (i.e.~bloom of green macroalgae of the genus \emph{Ulva}) were
reported in Brittany, France, back in the 1970s and have since been a
concern for local policies and economic activities
\citep{menesguen2018marees}. Some regions of the world have witnessed an
increase in brown macroalgae blooms, predominantly involving algae of
the genus \emph{Sargassum} washing along the Caribbean coastlines
\citep{louime2017sargassum}, and more recently \emph{Rugulopteryx
okamurea} in southern Europe \citep{Roca2022}. Satellite remote sensing
has proven to be a valuable tool for mapping the spatial and temporal
extent of macroalgal blooms worldwide. However, due to limitations in
spatial resolution, it can only effectively map well-developed blooms
\citetext{\citealp[ ]{rs13081408}; \citealp[
]{klemas2012remote}; \citealp{haro2023biointertidal}}. High spatial
resolution drone imagery, coupled with an accurate classification
algorithm, could be used to map the early stages of macroalgal blooms in
areas known to have regular blooms or in new sites. Indeed, this
approach could provide early warning alerts to local managers.

Employing traditional sampling methods to monitor coastal ecosystems is
time and resource-intensive, and the findings are often difficult to
scale-up. Earth Observation can bridge this gap and meet the needs for
systematic monitoring coastal ecosystems over large areas
\citep{papathanasopoulou2019satellite}. The retrieval of Essential
Biodiversity Variables and Essential Ocean Variables through satellite
observations has been increasingly common, enabling comprehensive
monitoring of entire ecosystems over extended time periods
\citetext{\citealp[ ]{ratnarajah2023monitoring}; \citealp{Zoffoli2021}}.
The Water Framework Directive \citep{WFD2000} mandates the achievement
and maintenance of ``good ecological status'' for all European waters,
which necessitates a comprehensive understanding and monitoring of
aquatic ecosystems, including coastal habitats like seagrass beds
\citetext{\citealp[ ]{foden2007angiosperms}; \citealp[
]{nordlund2024one}; \citealp{Zoffoli2021}}.
